<div align="center"> 
### Impact of Severe Weather Events 
### on Public Health and Economy of the United States

Amy Jiang    
November 21 2014  </div>
___
  
#### Problem Statement
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis:

- Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
- Across the United States, which types of events have the greatest economic consequences?
<br /><br />

#### Basic settings
```{r}
echo = TRUE  # Always make code visible
library(ggplot2)
library(plyr)
library(R.utils)
```
<br />

#### Data Processing
_The NOAA storm data file is downloaded from internet and unzipped to current folder._
```{r}
if (!"stormData.csv.bz2" %in% dir("./")) {
    download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "stormData.csv.bz2")
}
if(!"stormData.csv" %in% dir("./")) {
    bunzip2("stormData.csv.bz2", "stormData.csv", overwrite=T, remove=F)
}
```
  
_The dimension, and content of NOAA data are studied._
```{r}
data.storm <- read.csv("stormData.csv", sep = ",")
dim(data.storm)
head(data.storm, n=3)
```
  
_A plot graph of storm events distributed years showing that the storm data is more complete in recent years._
```{r}
# Since we will only care about the year of the event date, format it to the number of year
data.storm$year <- as.numeric(format(as.Date(data.storm$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
# A quick histogram to see how data was distributed across the years
hist(data.storm$year, xlab= "Year", breaks = 30, main = "Distribution of Weather Events Data over Years", col="steel blue")
```
  
Following the histogram in above, we could see that the weather events captured since 1995 have been almost doubled than in the past.  We believe this increase in the number of captured events is due to the advancement of modern technlogy, which enabled scientists to record the event that they were not able to capture before.  Since our study is focused on the total impact of weather events over years, I have chosen to not discriminate data based on this known observance.  <br /><br />

#### Study of Severe Weather Events' Impact on Public Health
_We are interested to see the top 15 types of weather events that resulted fatalities and injuries._  

_Top 15 types of weather events that have the highest fatality rate:_
```{r}
# Subtotal of fatablities by event type
data.storm.fatality <- aggregate(data.storm[, "FATALITIES"], by = list(data.storm$EVTYPE), FUN = "sum", na.rm=T)
# Sorting the data by numbers of fatalities in descending order
data.storm.fatality  <- arrange(data.storm.fatality, data.storm.fatality[,2], decreasing=T)
# Subsetting the top 15
data.storm.fatality <- data.storm.fatality[1:15,]
# Adding column headers
colnames(data.storm.fatality)  <- c("Event.Type", "Fatalities")
# Setting levels of Event.Type
data.storm.fatality[,"Event.Type"]  <- factor(data.storm.fatality[,"Event.Type"], levels = data.storm.fatality[,"Event.Type"])

# Show the top 15 fatality data
data.storm.fatality

# Plotting the figure
ggplot(data.storm.fatality, aes(Event.Type, Fatalities)) + geom_bar(stat = "identity", color="steel blue", fill="steel blue", width = 0.7) + theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + labs(x="Event Type", y="Total No. of Fatalities", title="Total Fatalities by Severe Weather\n Events in the U.S.\n from 1950 - 2011\n")
```

_Top 15 types of weather events that have the highest injury rate:_
```{r}
# Subtotal of injuries by event type
data.storm.injury <- aggregate(data.storm[, "INJURIES"], by = list(data.storm$EVTYPE), FUN = "sum", na.rm=T)
# Sorting the data by numbers of injuries in descending order
data.storm.injury  <- arrange(data.storm.injury, data.storm.injury[,2], decreasing=T)
# Subsetting the top 15
data.storm.injury <- data.storm.injury[1:15,]
# Adding column headers
colnames(data.storm.injury)  <- c("Event.Type", "Injuries")
# Setting levels of Event.Type
data.storm.injury[,"Event.Type"]  <- factor(data.storm.injury[,"Event.Type"], levels = data.storm.injury[,"Event.Type"])

# Show the top 15 injury data
data.storm.injury

# Plotting the figure
ggplot(data.storm.injury, aes(Event.Type, Injuries)) + geom_bar(stat = "identity", color="steel blue", fill="steel blue", width = 0.7) + theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + labs(x="Event Type", y="Total No. of Injuries", title="Total Injuries by Severe Weather\n Events in the U.S.\n from 1950 - 2011\n")
```
<br />

#### Study of Severe Weather Events' Impact on Economy
_We are interested to see the top 15 types of weeather events that resulted loss in property damage and crop damage._   

_Top 15 types of weather events that have caused the highest property damage:_
```{r}
# Remove unnecessary columns from data
data.storm.property.damage <- subset(data.storm, select = c("EVTYPE", "PROPDMG", "PROPDMGEXP"))
# Transform the measurements to number
data.storm.property.damage <- mutate(data.storm.property.damage, 
                      PROPDMGTTL = ifelse(PROPDMGEXP=="k" | PROPDMGEXP=="K", 
                                          PROPDMG * 1000, 
                                          ifelse(PROPDMGEXP=="m" | PROPDMGEXP=="M",
                                                 PROPDMG * 1000000,
                                                 ifelse(PROPDMGEXP=="b" | PROPDMGEXP=="B",
                                                        PROPDMG * 1000000000,
                                                        ifelse(PROPDMGEXP=="h" | PROPDMGEXP=="H",
                                                                PROPDMG * 100, PROPDMG)))))
# Calculating total by event types
data.storm.property.damage <- aggregate(data.storm.property.damage$PROPDMGTTL, by = list(data.storm.property.damage$EVTYPE), FUN = "sum", na.rm=T)
# Sort
data.storm.property.damage <- arrange(data.storm.property.damage, data.storm.property.damage[,2], decreasing=T)
# Select top 15
data.storm.property.damage <- data.storm.property.damage[1:15,]
# Adding column names
colnames(data.storm.property.damage)  <- c("Event.Type", "Property.Damage.Total")
# Setting levels of Event Type
data.storm.property.damage[,"Event.Type"]  <- factor(data.storm.property.damage[,"Event.Type"], levels = data.storm.property.damage[,"Event.Type"])

# Display the top 15 property damage table
data.storm.property.damage
# Plotting the figure
ggplot(data.storm.property.damage, aes(Event.Type, Property.Damage.Total/1e9)) + geom_bar(stat = "identity", color="steel blue", fill="steel blue", width = 0.7) + theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + labs(x="Event Type", y="Property Damage (in billion dollars)", title="Total Property Damages by Severe Weather\n Events in the U.S.\n from 1950 - 2011\n")
```

_Top 15 types of weather events that have caused the highest crop damage:_
```{r}
# Remove unnecessary columns from data
data.storm.crop.damage <- subset(data.storm, select = c("EVTYPE", "CROPDMG", "CROPDMGEXP"))
# Transform the measurements to number
data.storm.crop.damage <- mutate(data.storm.crop.damage, 
                      CROPDMGTTL = ifelse(CROPDMGEXP=="k" | CROPDMGEXP=="K", 
                                          CROPDMG * 1000, 
                                          ifelse(CROPDMGEXP=="m" | CROPDMGEXP=="M",
                                                 CROPDMG * 1000000,
                                                 ifelse(CROPDMGEXP=="b" | CROPDMGEXP=="B",
                                                        CROPDMG * 1000000000,
                                                        ifelse(CROPDMGEXP=="h" | CROPDMGEXP=="H",
                                                                CROPDMG * 100, CROPDMG)))))
# Calculating total by event types
data.storm.crop.damage <- aggregate(data.storm.crop.damage$CROPDMGTTL, by = list(data.storm.crop.damage$EVTYPE), FUN = "sum", na.rm=T)
# Sort
data.storm.crop.damage <- arrange(data.storm.crop.damage, data.storm.crop.damage[,2], decreasing=T)
# Select top 15
data.storm.crop.damage <- data.storm.crop.damage[1:15,]
# Adding column names
colnames(data.storm.crop.damage)  <- c("Event.Type", "Crop.Damage.Total")
# Setting levels of Event Type
data.storm.crop.damage[,"Event.Type"]  <- factor(data.storm.crop.damage[,"Event.Type"], levels = data.storm.crop.damage[,"Event.Type"])

# Display the top 15 property damage table
data.storm.crop.damage
# Plotting the figure
ggplot(data.storm.crop.damage, aes(Event.Type, Crop.Damage.Total/1e9)) + geom_bar(stat = "identity", color="steel blue", fill="steel blue", width = 0.7) + theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + labs(x="Event Type", y="Crop Damage (in billion dollars)", title="Total Crop Damages by Severe Weather\n Events in the U.S.\n from 1950 - 2011\n")
```
<br />

#### Conclusion
Following our analysis on the provided data, Tornado and Excessive Heat are the top two severe weather events that have been most harmful to public health, while Drought and Flood are the top two severe weather events that have caused the greatest economic loss in the United States.
<br />